---
title: "Class 12"
author: "Izabelle Querubin"
format: pdf
---

## 1. Bioconductor and DESeq2 setup

```{r}
library(BiocManager)
library(DESeq2)
```

# 2. Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <- read.csv("airway_metadata.csv")
```

```{r}
nrow(counts)
```

```{r}
View(metadata)
```

### **Q1.** How many genes are in this dataset?

38694 genes

### **Q2.** How many \'control\' cell lines do we have?

4 'control' cell lines

# 3. Toy differential gene expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

### Q3. How would you make the above code in either approach more robust?

```{r}
# Subset metadata based on "dex" condition
control <- metadata[metadata$dex == "control", ]

# Validate if "id" column exists in control
if (!"id" %in% colnames(control)) stop("Control metadata does not contain 'id' column.")

# Subset counts based on control$id
control.counts <- counts[, control$id]

# Calculate control mean
control.mean <- rowSums(control.counts) / 4
head(control.mean)
```

This establishes a "validation check" to ensure that the "id" column exists in control.

### Q4. Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

```{r}
# Subset metadata based on "dex" condition
treated <- metadata[metadata[,"dex"] == "treated",]

# Validate if "id" column exists in treated
if (!"id" %in% colnames(treated)) stop("Treated metadata does not contain 'id' column.")

# Calculate treated mean
treated.mean <- rowSums(counts[, treated$id]) / 4 
names(treated.mean) <- counts$ensgene
head(treated.mean)
```

```{r}
meancounts <- data.frame(Control = control.mean, Treated = treated.mean)
colSums(meancounts)
```

### Q5(a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples. 

```{r}
# Create a scatter plot using base R
plot(control.mean, treated.mean, 
     xlab = "Mean of Control Samples", 
     ylab = "Mean of Treated Samples",
     main = "Mean of Treated Samples vs. Mean of Control Samples")
```

### **Q5 (b).**You could also use the **ggplot2** package to make this figure producing the plot below. What **geom\_?()** function would you use for this plot?

```{r}
library(ggplot2)

# Create the scatter plot using ggplot2
ggplot(meancounts, aes(x = Control, y = Treated)) +
  geom_point() +
  labs(x = "Mean of Control Samples", y = "Mean of Treated Samples") +
  ggtitle("Mean of Treated Samples vs. Mean of Control Samples")
```

geom_point()

### **Q6.** Try plotting both axes on a log scale. What is the argument to **plot()** that allows you to do this?

```{r}
plot(control.mean, treated.mean,
     xlab = "Mean of Control Samples",
     ylab = "Mean of Treated Samples",
     main = "Mean of Treated Samples vs. Mean of Control Samples",
     log = "xy")
```

log = "xy"

```{r}
meancounts$log2fc <- log2(meancounts[,"Treated"]/meancounts[,"Control"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

### Q7. What is the purpose of the `arr.ind` argument in the **which()** function call above? Why would we then take the first column of the output and need to call the **unique()** function?

The **`arr.ind = TRUE`** argument tells **`which()`** to return the results as indices in array/matrix form, allowing us to extract the row indices of the elements with zero values.

The first column of the output is extracted because it contains the row indices of elements with zero values. **`unique()`** is used to obtain only the unique row indices since there might be cases where multiple elements have a value of zero in the same row.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

### Q8. Using the `up.ind` vector above can you determine how many up regulated genes we have at the greater than 2 fc level?

```{r}
sum(up.ind)
```

250 up regulated genes.

### Q9. Using the `down.ind` vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

```{r}
sum(down.ind)
```

367 down regulated genes.

### Q10. Do you trust these results? Why or why not?

The log2 fold change threshold of greater than 2 or less than -2 is a common cutoff used in differential expression analysis. Because of this, I think we are on the right path but more statistical analysis (specifically significance) is needed to ensure that these results are not misleading.

# 4. DESeq2 analysis

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

```{r}
# Run the DESeq pipeline on the dds object, and reassign the whole thing back to dds
dds <- DESeq(dds)

# Get results
res <- results(dds)
res
```

```{r}
res.df <- as.data.frame(res)
View(res)
```

```{r}
summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# 5. Adding annotation data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),      # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
head(res)
```

### Q11. Run the **mapIds()** function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and `res$genename`.

```{r}
# Add Entrez ID column
res$entrez <- mapIds(org.Hs.eg.db,
                     keys = row.names(res),
                     keytype = "ENSEMBL",
                     column = "ENTREZID",
                     multiVals = "first")

# Add UniProt accession column
res$uniprot <- mapIds(org.Hs.eg.db,
                      keys = row.names(res),
                      keytype = "ENSEMBL",
                      column = "UNIPROT",
                      multiVals = "first")

# Add gene name column
res$genename <- mapIds(org.Hs.eg.db,
                       keys = row.names(res),
                       keytype = "ENSEMBL",
                       column = "GENENAME",
                       multiVals = "first")
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

# 6. Data Visualization

```{r}
plot( res$log2FoldChange, -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)

```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```
