---
title: "Class 17"
author: "Izabelle Querubin"
format: pdf
---

# COVID-19 Vaccination Rates

## Getting Started

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

### Q1. What column details the total number of people fully vaccinated?

"persons_fully_vaccinated"

### Q2. What column details the Zip code tabulation area?

"zip_code_tabulation_area"

```{r}
min(vax$as_of_date)
```

### Q3. What is the earliest date in this dataset?

2021-01-05

```{r}
max(vax$as_of_date)
```

### Q4. What is the latest date in this dataset?

2023-06-13

```{r}
skimr::skim_without_charts(vax)
```

### Q5. How many numeric columns are in this dataset?

14

```{r}
num_na <- sum(is.na(vax$persons_fully_vaccinated))
```

### Q6. Note that there are "missing values" in the dataset. How many `NA` values there in the `persons_fully_vaccinated` column?

18076

ncol(vax\$persons_fully_vaccinated, na.rm = TRUE)

```{r}
percent_missing <- round((num_na / nrow(vax)) * 100, 2)
percent_missing
```

### Q7. What percent of `persons_fully_vaccinated` values are missing (to 2 significant figures)?

8.01%

```{r}
library(lubridate)
```

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[1]
```

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

### **Q9.** How many days have passed since the last update of the dataset?

3 days have passed

### **Q10.** How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
num_unique_dates <- length(unique(vax$as_of_date))
num_unique_dates
```

128 unique dates

## Working with Zip Codes

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )
```

```{r}
# Pull data for all ZIP codes in the dataset
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```

# Focus on the San Diego area

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
head(sd.10)
```

```{r}
distinct_codes <- unique(sd$zip)
distinct_zip_codes <- length(distinct_codes)
distinct_zip_codes
```

### **Q11.** How many distinct zip codes are listed for San Diego County?

107

```{r}
largest_population <- sd$zip[which.max(sd$tot_population)]
largest_population
```

### **Q12.** What San Diego County Zip code area has the largest population in this dataset?

92154

```{r}
library(dplyr)

sd_filtered <- filter(vax, county == "San Diego" & as_of_date == "2023-05-23")
```

```{r}
sd_filtered <- mutate(sd_filtered, percent_of_population_fully_vaccinated = as.numeric(percent_of_population_fully_vaccinated))
```

```{r}
average_percent_vaccinated <- mean(sd_filtered$percent_of_population_fully_vaccinated, na.rm = TRUE)
average_percent_vaccinated <- round(average_percent_vaccinated, 4)
average_percent_vaccinated
```

### **Q13.** What is the overall average (with 2 decimal numbers) "Percent of Population Fully Vaccinated" value for all San Diego "County" as of "2023-05-23"?

74.21%

### Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of "2023-05-23"?

```{r}
hist(sd_filtered$percent_of_population_fully_vaccinated, breaks = 10, col = "skyblue", border = "black",
     xlab = "Percent of Population Fully Vaccinated", ylab = "Frequency",
     main = "Distribution of Percent of Population Fully Vaccinated")
```

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

### Q15. Using **ggplot** make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
library(ggplot2)

ggplot(ucsd) +
  aes(x = as.Date(as_of_date), y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0, 1)) +
  labs(x = "Date", y = "Percent Vaccinated", title = "Vaccination Rate for La Jolla 92037")
```

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-05-23")

head(vax.36)
```

### Q16. Calculate the mean *"Percent of Population Fully Vaccinated"* for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23". Add this as a straight horizontal line to your plot from above with the `geom_hline()` function?

```{r}
mean_percent_vaccinated <- mean(ucsd$percent_of_population_fully_vaccinated, na.rm = TRUE)

ggplot(ucsd) +
  aes(x = as.Date(as_of_date), y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group = 1) +
  ylim(c(0, 1)) +
  labs(x = "Date", y = "Percent Vaccinated", title = "Vaccination Rate for La Jolla 92037") +
  geom_hline(yintercept = mean_percent_vaccinated, linetype = "dashed", color = "red")
```

### Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the *"Percent of Population Fully Vaccinated"* values for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* "2023-05-23"?

```{r}
summary_stats <- summary(ucsd$percent_of_population_fully_vaccinated)
summary_stats
```

### Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36, aes(x = percent_of_population_fully_vaccinated)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "blue", alpha = 0.5) +
  labs(x = "Percent Vaccinated", y = "Count", title = "Histogram of Vaccination Rate") +
  theme_minimal()
```

### Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

```{r}
vax %>% filter(as_of_date == "2023-05-23") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```

The 92109 and 92040 zip code areas are below the average value.

### Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a `age5_plus_population > 36144`.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(x = as.Date(as_of_date), y = percent_of_population_fully_vaccinated, group = zip_code_tabulation_area) +
  geom_line(alpha = 0.2, color = "blue") +
  ylim(0, 1) +
  labs(x = "Date", y = "Percent of Population Fully Vaccinated",
       title = "Vaccination Progress by ZIP Code",
       subtitle = "Areas with population > 36144") +
  geom_hline(yintercept = mean(vax.36.all$percent_of_population_fully_vaccinated), linetype = "dashed") 
```
