---
title: "class13"
author: "Izabelle Querubin"
format: pdf
---

# Class 13 Mini-Project

## Section 1. Differential Expression Analysis

```{r}
library(DESeq2)
```

```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peak
colData = read.csv(metaFile, row.names=1)
head(colData)
```

```{r}
# Import countdata
countData = read.csv(countFile, row.names=1)
head(countData)
```

#### **Q1**. Complete the code below to remove the troublesome first column from `countData` .

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[, -1])
head(countData)
```

#### **Q2**. Complete the code below to filter `countData` to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).

```{r}
# Filter count data where you have 0 read count across all samples.
countData = countData[rowSums(countData) > 0, ]
head(countData)
```

### Running DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
```

```{r}
dds = DESeq(dds)
```

```{r}
dds
```

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

```{r}
summary(res)
```

### Volcano Plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

#### **Q4**. Improve this plot by completing the below code, which adds color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

### Adding gene annotation

#### **Q5**. Use the **mapIDs()** function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

#### **Q6**. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```

## Section 2. Pathway Analysis

### KEGG Pathways

```{r}
library(pathview)
```

```{r}
library(gage)
```

```{r}
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
# Look at the first few down (less) pathways
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

```{r}
# Extract the top 5 down-regulated pathways
keggresdownpathways <- rownames(keggres$less)[1:5]

# Extract the 8-character long IDs part of each string
keggresdownids <- substr(keggresdownpathways, start = 1, stop = 8)

# Plot the pathview figures for the top 5 down-regulated pathways
pathview(gene.data = foldchanges, pathway.id = keggresdownids, species = "hsa")
```

## Section 3. Gene Ontology (GO)

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

## Section 4. Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

#### **Q8**: What pathway has the most significant \"Entities p-value\"? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The "Cell Cycle, Mitotic" has the most significant "Entities p-value". While my previous KEGG results include DNA Replication, Nucleocytoplasmic Transport, Homologous Recombination, etc., the Reactome Analysis' most significant pathways include Cell Cycle (Mitotic), Cell Cycle, Mitotic Spindle Checkpoint, Amplification of signal from unattached kinetochores via a MAD2 inhibitory signal, etc. These differences could be caused by the database content. These are two different databases, which suggests that some pathways or genes may be more unique to one database than the other. These differences could also be caused by the statistical algorithms each database may possess. KEGG and Reactome could employ different statistical approaches to their data which could lead to variances in p-values and significance levels.
